---
title: "Specifying models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specifying models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bayesplay)
```

The core `bayesplay` workflow is:

1. Specify a likelihood for the observed data.
2. Specify a prior for each model.
3. Multiply each prior by the likelihood.
4. Integrate each product to obtain a marginal likelihood.
5. Divide the marginal likelihoods to obtain a Bayes factor.

This vignette is a compact reference for specifying those likelihoods and
priors.

## Likelihood families

The `likelihood()` function takes a `family` argument plus the parameters needed
for that family.

```{r, echo = FALSE}
knitr::kable(data.frame(
  Family = c(
    "normal",
    "student_t",
    "binomial",
    "noncentral_t",
    "noncentral_d",
    "noncentral_d2"
  ),
  Use = c(
    "A normally distributed estimate, such as a mean difference and standard error.",
    "A scaled and shifted t likelihood.",
    "A count of successes from a fixed number of trials.",
    "A t statistic and degrees of freedom.",
    "A one-sample or paired-samples Cohen's d and sample size.",
    "An independent-samples Cohen's d and two group sample sizes."
  ),
  Required_parameters = c(
    "mean, sd",
    "mean, sd, df",
    "successes, trials",
    "t, df",
    "d, n",
    "d, n1, n2"
  ),
  check.names = FALSE
))
```

For example, a normal likelihood can represent an observed mean difference and
its standard error:

```{r}
normal_likelihood <- likelihood(family = "normal", mean = 5.5, sd = 32.35)
normal_likelihood
```

A binomial likelihood can represent a number of successes out of a fixed number
of trials:

```{r}
binomial_likelihood <- likelihood(family = "binomial", successes = 7, trials = 10)
binomial_likelihood
```

## Prior families

The `prior()` function uses the same `family` pattern.

```{r, echo = FALSE}
knitr::kable(data.frame(
  Family = c("point", "normal", "student_t", "cauchy", "uniform", "beta"),
  Use = c(
    "A model that places all prior mass on one parameter value.",
    "A normal prior, optionally truncated with range.",
    "A scaled and shifted t prior, optionally truncated with range.",
    "A Cauchy prior, optionally truncated with range.",
    "A flat prior between lower and upper bounds.",
    "A prior over probabilities from 0 to 1."
  ),
  Required_parameters = c(
    "point",
    "mean, sd",
    "mean, sd, df",
    "location, scale",
    "min, max",
    "alpha, beta"
  ),
  check.names = FALSE
))
```

A point prior is commonly used for a null model:

```{r}
null_prior <- prior(family = "point", point = 0)
null_prior
```

Continuous priors can be truncated with `range`. For example, this half-normal
prior assigns density only to positive parameter values:

```{r}
positive_prior <- prior(family = "normal", mean = 0, sd = 13.3, range = c(0, Inf))
positive_prior
```

For binomial likelihoods, a beta prior is often convenient because it is defined
on the same 0 to 1 parameter space:

```{r}
beta_prior <- prior(family = "beta", alpha = 2.5, beta = 3.5)
beta_prior
```

## Calculating a Bayes factor

The order of the division determines the interpretation. In the example below,
`bf_10` is the Bayes factor for the alternative model over the null model. A
value larger than 1 favours the alternative model, and a value smaller than 1
favours the null model.

```{r}
data_model <- likelihood(family = "normal", mean = 5.5, sd = 32.35)
alternative_prior <- prior(family = "normal", mean = 0, sd = 13.3, range = c(0, Inf))
null_prior <- prior(family = "point", point = 0)

m1 <- integral(data_model * alternative_prior)
m0 <- integral(data_model * null_prior)

bf_10 <- m1 / m0
summary(bf_10)
```

Reversing the division gives the reciprocal Bayes factor:

```{r}
bf_01 <- m0 / m1
summary(bf_01)
```

